Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-10 write-ups/PLDI/paper.tex
Structure.

The situation is more stable if a cooperative kernel does not directly
issue $\offerkill$ and $\offerfork$ calls directly.

 then the only possibility for the number of
workgroups to change is due to $\resizingglobalbarrier$ being invoked.

\ADComment{Shorten sentences.}
In this case, because the resizing barrier is a kernel-level function,
the statements executed by the cooperative kernel can be divided into
\emph{resizing barrier intervals}: a resizing barrier interval starts
at a resizing barrier or the kernel entry point, and ends at a
resizing barrier or a kernel exit point.  At any moment of execution,
all threads are guaranteed to be executing within the same resizing
barrier interval.  As a result, the number of workgroups executing the
kernel is guaranteed to remain unchanged within each resizing barrier
interval, so the result returned by $\getnumgroups{0}$ is guaranteed
to be identical on repeated calls during such an interval.

This guarantee may be exploited by algorithms that perform strided data
processing.  \ADComment{Tyler, can we give an example of a data
  processing loop that would have to be changed?}

@@

\ADComment{TODO: needs work.}
.  The irregular
algorithms we encountered either employ work stealing or operate on
graph data structures.  Work stealing algorithms have a transactional
flavour, whereby tasks are repeatedly pulled from a queue and
processed by a workgroup.  The point of task completion---the end of a transaction---is a natural place for a workgroup to invoke both $\offerkill$ and $\offerfork$, to indicate that

@@

The semantics presented in Section~\ref{sec:semantics} describe the envelope of
behaviour that a developer of a cooperative kernel should be prepared
for.

However, the aim of cooperative kernels is to find a balance that
allows \emph{efficient execution} of long-running kernels
(e.g.\ implementing irregular data-parallel algorithms), and
\emph{responsive multi-tasking} so that the GPU can be shared between
one or more cooperative kernels and a number of short, high-priority
tasks with soft real-time constraints.

@@

\ADComment{Among other things, make sure we address the point that placing the primitives was pretty easy.}

@@

\[
\inferrule{
\forall j \;.\; w_M(j) = (l_j, \offerkill();\mathit{ss})
}
{
(\sigma, (\dots, w_{M-1}, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_{M-1}, \bot, \bot, \dots, \bot))
}
(\textsc{Kill})
\]

@@

\[
\inferrule{
\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \resizingglobalbarrier();\mathit{ss})
\\
\forall i \neq 0 \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \globalbarrier(); \globalbarrier(); \offerkill(); \globalbarrier();\mathit{ss})
\\
\forall j\;.\;w_0'(j) = (l_{0,j}, \globalbarrier(); \offerfork(); \globalbarrier(); \offerkill(); \globalbarrier();\mathit{ss})
}
{
(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (w_1', \dots, w_M', \bot, \dots, \bot))
}
(\textsc{Resizing-Barrier})
\]

>>>>

The situation is more stable if a cooperative kernel does not directly
issue $\offerkill$ and $\offerfork$ calls directly.  In this case,
only calls to $\resizingglobalbarrier$ can cause the number of active
workgroups to change and, as discussed in Section~\ref{sec:semantics},
this function causes synchronisation between all threads and causes
the number of active workgroups to change atomically.  At any point
during execution, the threads of a kernel are executing between some
pair of resizing barrier calls, which we call a \emph{resizing barrier
  intervals} (here we consider the kernel entry and exit points
conceptually to be special cases of resizing barriers).  The number of
workgroups executing the kernel is thus constant within each resizing
barrier interval, i.e.\ the result returned by repeated calls to
$\getnumgroups()$ will be identical during such an interval.

This guarantee can be exploited by algorithms that perform strided
data processing.  \ADComment{Tyler, can we give an example of a data
  processing loop that would have to be changed?}

@@

It is up to the programmmer to decide where to place our new language
constructs.  This must be done bearing in mind both the semantics of
the constructs, as detailed in Section~\ref{ }, as well as the
non-functional requirements of cooperative kernels that we describe in
Section~\ref{sec:nonfunctional}.

We note, however, that the programming model and associated language
constructs were inspired by the requirements of the irregular GPU
algorithms proposed in the literature, and that at least for these
algorithms it is usually easy to identify where to place the
constructs.

The irregular algorithms we encountered either employ work stealing or
operate on graph data structures.

Work stealing algorithms have a transactional flavour, whereby tasks
are repeatedly pulled from a queue and processed by a workgroup.  The
point of task completion---the end of a transaction---is a natural
place for a workgroup to invoke both $\offerkill$ and $\offerfork$, to
indicate both that the workgroup could gracefully exit computation,
but also that there is more work to be processed and that further
workgroups would be useful in order to process this work.  Such
algorithms usually require little state to persist between tasks;
typically only the pointers to the queue data structures need to be
pre-initialised, by annotating them with $\transmit$.
\ADComment{Refer back to cooperative versions of examples in
  Section~\ref{sec:semantics}.}

The level-by-level approach employed by many graph algorithms tends to
involve global barrier synchronisation.  In some cases, a global
barrier is required in the midst of a transaction, in which case
$\globalbarrier$, but not $\resizingglobalbarrier$, is suitable for
achieving synchronization without disturbing the set of active
workgroups.  But often it is the case that on completing a level of
the graph algorithm, the next level could be processed by more or
fewer workgroups, something which $\resizingglobalbarrier$
facilitates.  Again, only a small amount of state is relevant to new
workgroups joining the computation, such as the value of the level
being processed; this relevant state must be annotated with
$\transmit$.

@@

The semantics presented in Section~\ref{sec:semantics} describe the envelope of
behaviour that a developer of a cooperative kernel should be prepared
for.
%
However, the aim of cooperative kernels is to find a balance that
allows \emph{efficient execution} of long-running kernels
(e.g.\ implementing irregular data-parallel algorithms), and
\emph{responsive multi-tasking} so that the GPU can be shared between
one or more cooperative kernels and a number of short, high-priority
tasks with soft real-time constraints.

@@

\ADComment{Describe the algorithms that we ported.}

\ADComment{Give stats on how many variables had to be marked with
  $\transmit$ (i.e., occurred in the restoration context), how many barriers we changed to resizing, how many offer kill and offer fork calls we included.}

\ADComment{Make sure we address the point that placing the primitives was pretty easy.}

\section{The GPUs we Tested}

\ADComment{Describe the platforms.  Explain why we could not try other
  platforms.}

\section{Determining Representative Workloads}

\ADComment{Better section title?}

\ADComment{Hugues, can you please explain your experimental
  methodology for working out required response times here?}

\paragraph{Experimental setup}  \ADComment{Describe the data we are gathering.}

\paragraph{Results}  \ADComment{Discuss the results.}

\section{The Overhad of Moving to Cooperative Kernels}

\ADComment{Results for stand-alone vs.\ cooperative without co-scheduling.}

\paragraph{Experimental setup}  \ADComment{Describe the data we are gathering.}

\paragraph{Results}  \ADComment{Discuss the results.}

\section{Performance and Responsiveness of Cooperative Scheduling}

\ADComment{Preamble}

\paragraph{Experimental setup}  \ADComment{Describe the data we are gathering.}

\paragraph{Results}  \ADComment{Discuss the results.}

@@

\[
\inferrule{
\forall j \;.\; w_M(j) = (l_j, \offerkill();\mathit{ss})
\\
$M > 0$
}
{
(\sigma, (\dots, w_{M-1}, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_{M-1}, \bot, \bot, \dots, \bot))
}
(\textsc{Kill})
\]

@@

\[
\inferrule{
\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \resizingglobalbarrier();\mathit{ss})
\\
\forall j\;.\;w_0'(j) = (l_{0,j}, \globalbarrier(); \offerfork(); \globalbarrier(); \globalbarrier();\mathit{ss})
\\
\forall i \neq 0 \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \globalbarrier(); \globalbarrier(); \offerkill(); \globalbarrier();\mathit{ss})
}
{
(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (w_1', \dots, w_M', \bot, \dots, \bot))
}
(\textsc{Resizing-Barrier})
\]
